FantasticLamp: a genome graph based pipeline for calculating the efficacy of genomic edits

### 2023-07-09 paper/main.tex
Edited small details

\address{
$^{\text{\sf 1}}$Department of Bioinformatics and Computational Biology, University of Stellenbosch, Stellenbosch, 7600, Western Cape, South Africa \\
$^{\text{\sf 2}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA \\
%
}

@@

Compared to linear alignment methods, genome graphs can provide a more accurate and comprehensive view \citep{garrison2018variation} of the relationships between sequences \citep{paten2017genome} because reads can be mapped to a graph that captures the variation between all the genomes/subsequences, instead of to a single reference. 
The cost of this increased accuracy and the ability to represent complexity is that using genome graphs is more computationally intensive than linear alignments methods \citep{rakocevic2019fast}.

@@

\section{Implementation}
\label{sec:implementation}
Here we describe the steps and tools used in creating \textit{FantasticLamp}. 
\subsection{Overview}

@@

The graph is converted into a format that is more efficient and finally indexed, both steps using \textit{vg} \citep{garrison2018variation}.
The pipeline then maps reads from the edited populations to this graph using \textit{vg}, creating a GAF (Gene Annotation Format) file representing the alignment.
A Python script is called to parse the alignment file and calculate the coverage of the homology arms compared to the reference homology arms.
The output of this pipeline is a coverage table in the form of a TSV file, which displays the name of each intended edit sequence along with the homology arm coverage and reference homology arm coverage.
A flow diagram of the steps and files used in the \textit{FantasticLamp} pipeline can be seen in .  
This provides a quantitative measure of the efficacy of each edit.

\subsection{Details}
Go into more detail. At the graph creation step, include a figure showing a locus with edits in Bandage with the counts (BandageNG works now).

Concluding paragraph.
This allows FantasticLamp to simultaneously quantify the efficacy of edits in multiple edited populations, which can be used to test novel editing methods as well as to verify current methods in experiments that involve genomic edits.
By using reads aligned to a graph instead of linear alignment, FantasticLamp can avoid reference bias, making it a useful tool for researchers in the field of genome engineering and other related fields who wish to precisely quantify genome edit states.\\

@@

\section{Discussion}
\label{sec:discussion}
As novel genome editing processes are developed in the near future, we posit that the need for software tools that can capture the complexity of the edits and the relationship between them will increase.
We show a basic approach for quantifying complex genome editing results, which are difficult to reliably and simply evaluate using a single reference genome that may lead to biased estimates of edit states.
The design library CSV file used in the creation of this pipeline was unique to a specific set of editing experiments performed at Inscripta Inc.
However, the basic format is generic to other editing experiments that users may carry out in the future, with only minimal changes to the script to adjust for difference in formatting of the design library file.
Future users will have to edit the bash script `find\_coverage.sh' such that the correct columns containing the edit- and reference homology arms are extracted.
While the pipeline was tested only on singleplex sequencing data, it is possible to perform analysis using pooled sequencing data, but data availability has limited our ability to thoroughly test this.
In summary, FantasticLamp provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits, and stands as a prototype for future work in this space.
\section*{Acknowledgements}
We are grateful to Inscripta Inc for providing the data from editing experiments that was used to create and test this pipeline. We thank Deanna Church for supporting our collaboration.

>>>>

\address{
$^{\text{\sf 1}}$Department of Bioinformatics and Computational Biology, University of Stellenbosch, Stellenbosch, 7600, Western Cape, South Africa \\
$^{\text{\sf 2}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA \\
$^{\text{\sf 3}}$Ian's affiliation goes here \\
%
}

@@

Compared to linear alignment methods, genome graphs can provide a more accurate and comprehensive view \citep{garrison2018variation} of the relationships between sequences \citep{paten2017genome} because reads can be mapped to a graph that captures the variation between all the genomes/subsequences, instead of to a single reference, as seen in Figure \ref{fig:images} (A), (B), and (C).
The cost of this increased accuracy and the ability to represent complexity is that using genome graphs is more computationally intensive than linear alignments methods \citep{rakocevic2019fast}.

@@

\section{Implementation}
\label{sec:implementation}
Here we describe the steps and tools used in creating \textit{FantasticLamp}.

@@

The graph is converted into a format that is more efficient and finally indexed, both steps using \textit{vg} \citep{garrison2018variation}.
The pipeline then maps reads from the edited populations to this graph using \textit{vg}, creating a GAF (Gene Annotation Format) file representing the alignment.
A Python script is called to parse the alignment file and calculate the coverage of the homology arms compared to the reference homology arms.
The output of this pipeline is a coverage table in the form of a TSV file, which displays the name of each intended edit sequence along with the homology arm coverage and reference homology arm coverage providing a quantitative measure of the efficacy of each edit.

This allows \textit{FantasticLamp} to simultaneously quantify the efficacy of edits in multiple edited populations, which can be used to test novel editing methods as well as to verify current methods in experiments that involve genomic edits.
By using reads aligned to a graph instead of linear alignment, \textit{FantasticLamp} can avoid reference bias, making it a useful tool for researchers in the field of genome engineering and other related fields who wish to precisely quantify genome edit states.\\
A flow diagram of the steps and files used in the \textit{FantasticLamp} pipeline can be seen in Figure \ref{fig:images}(D).

@@

\section{Discussion}
\label{sec:discussion}
As novel genome editing processes are developed in the near future, we posit that the need for software tools that can capture the complexity of the edits and the relationship between them will increase.
We show a basic approach for quantifying complex genome editing results, which are difficult to reliably and simply evaluate using a single reference genome that may lead to biased estimates of edit states.
The design library CSV file used in the creation of this pipeline was unique to a specific set of editing experiments performed at Inscripta Inc.
However, the basic format is generic to other editing experiments that users may carry out in the future, with only minimal changes to the script to adjust for difference in formatting of the design library file.
Future users will have to edit the bash script `find\_coverage.sh' such that the correct columns containing the edit- and reference homology arms are extracted.
While the pipeline was tested on artificial data and singleplex sequencing data, it is possible to perform analysis using pooled sequencing data, but data availability has limited our ability to thoroughly test this.
In summary, \textit{FantasticLamp} provides a basic demonstration of the principle of using a variation graph as a reference system to avoid bias when quantifying genomic edits, and stands as a prototype for future work in this space.
\section*{Acknowledgements}
We are grateful to Inscripta Inc for providing the data from editing experiments that was used to create and test this pipeline. We thank Deanna Church for supporting our collaboration.
